A Large-scale Study of Polymorphism in R

### 2019-04-02 paper/main.tex

L0 is limited in expressivity. To address its limitations, we augment it with
scalars and parametrized lists.  In addition, we collect attribute and class
information to distinguish a type for data frames and matrices, as the
\code{typeof} these is simply {\tt list} in R.

The second level enhances {\bf L0} with scalar values for primitive types
(\sI, \sD, \sC, \sL, \sR, and \sX), type parametric lists (written \lT{T}
for any type {\tt T}), data frames \attr{\l}{df}, and matrices
\attr{\vec{\tt T}}{mat}.

@@

Though more expressive than L0, L1 only incorporates a limited amount of
class and attribute information. For instance, L1 says nothing about user-defined
classes and non-standard attributes. The standard attribute patterns in R are:

\begin{itemize}
\item data frames have three attributes: {\tt names} for column names, {\tt
  row.names} for row names, and {\tt class} (which is instantiated to {\tt
  "data.frame"} upon creation).
\item matrices have at least a {\tt dim} attribute with matrix dimensions,
  and optionally a {\tt dimnames} attribute for dimension names. Curiously,
  the {\tt dim} attribute is all that is needed to define a matrix (i.e., a
  vector with a {\tt dim} attribute {\it is} a matrix).
\item lists and vectors may have a {\tt names} attribute, which assigns
  names to locations.
\item to ease dealing with large sorted sets of non-numeric values, R offers
  the \code{factor} and \code{levels} functions.  \code{factor(x,
    levels=someOrder)} adds a {\tt "levels"} attribute to \code{x}, which
  specifies an ordering for \code{x} according to the \code{sortedOrder}
  parameter.  \code{factor} also changes the class of the parameter being
  factored to \texttt{"factor"}.
\end{itemize}

@@

\begin{table}[ht]\label{tab:L0argcounts}\centering\begin{tabular}{lrr}  \hline
Type                   & Count & Percentage \\  \hline
%  Full Monomorphic & 785K & 85.87 \\ 
  Monomorphic in Type & 870K & 95.2\% \\ 
%  Monomorphic in Class & 827K & 90.41 \\ 
%  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
  Total Seen & 914K & --- \\ 
   \hline
\end{tabular}
\caption{Argument position polymorphism in L0.}
\end{table}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsL1}
\centering
\begin{tabular}{lrr}
  \hline
 Function Polymorphism & Count & Percentage \\
  \hline
%  Full Monomorphic & 143K & 65.88 \\
  Monomorphic in Type & 167K & 77.25 \\
%  Monomorphic in Class & 163K & 75.24 \\
%  Monomorphic in Attribute Pattern & 161K & 74.37 \\
  Total Seen & 216K & --- \\
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in L1.}
\end{table}

>>>>

{\bf L0} is limited in expressivity as it supports polymoprhism for list in
a trivial way, it does not address some of the popular constructed data
types such as matrices and data frames, and cannot differentiate between
scalar values and vectors.  {\bf L1} types build on and extend the previous
level with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR,
and \sX), type parametric lists (written \lT{T} for any type {\tt T}), data
frames (written \attr{\l}{df} as they are built up from lists of type any),
and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
Subtyping is extended such that {\tt T}<:{$\vec{\tt T}$} for any primitive
type {\tt T}. There is no subtyping relation between lists of different
types.

@@

{\bf L2} extends the previous system with more support for attributes.
While there is a finite number of results from the {\tt typeof} function,
this is not the case for \code{class} and \code{attributes}.  Classes and
attributes are user-defined.  Of course, not all classes are user-defined,
and R has a number of built-in or otherwise standard classes, ranging from
truly primitive classes such as {\tt numeric} to the more complex {\tt
  data.frame}.  We have already used the latter in {\bf L1}. We will ignore
those classes and thus highlight user-defined ones.  As for attributes,
again some patterns arise naturally, for instance all matrices have a {\tt
  dims} attribute (indeed, it's this attribute that makes a matrix a
matrix).  Attributes are name-value pairs, to keep things simple we reduce
these to names. The only attribute that we treat specially is the {\tt
  class} for those we will recurse and capture the set of values defined by
the user (these are the class names for the object).  Thus, we extend {\bf
  L1} with attribute lists and classes: \attrclass{T}{C}{N} where {\tt T}
can be any L1 type other than data frame or matrix, {\tt C} and {\tt N}
stands for a list names.  The subtype relation is extended with
\attrclass{T}{C}{N}<:\attrclass{T}{C'}{D'} if {\tt C} $\subseteq$ {\tt C'}
and {\tt N}$\subseteq${\tt N'}.

@@

\begin{table}[ht]\label{tab:L0argcounts}\centering\begin{tabular}{lrr}  \hline
Type                   & Count & Percentage \\  \hline
  Monomorphic in Type & 845K & 92.46 \\
  Total Seen & 914K & --- \\
   \hline
\end{tabular}
\caption{Argument position polymorphism in L0.}
\end{table}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsL1}
\centering
\begin{tabular}{lrr}
  \hline
 Function Polymorphism & Count & Percentage \\
  \hline
  Monomorphic in Type & 167K & 77.25 \\
  Total Seen & 216K & --- \\
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in L1.}
\end{table}
